### 2012-06-20 trunk/ijcis-pos-model/sources/temporaldatabases.tex
version sent to Olga and Nicolas
final draft

In this paper we will consider two kinds of imperfection:
\begin{itemize}
\item \textbf{Imperfection in the database} the knowledge about the temporal data contains some imperfection. E.g., a database record shows that \emph{`The car was in the garage around April.'}
 \item \textbf{Imprecision in the query specification} denotes the imprecision in the specification of temporal criteria by the user, when querying. E.g., \emph{`The user wants a car which was in the garage around April.'}
\end{itemize}

@@

In order to deal with uncertainty in time intervals, several proposals have been made. Here, two approaches are introduced: the first one, based on \emph{Fuzzy Validity Periods}~\cite{Garrido2009} and the second one, based on \emph{Possibilistic Valid-time Periods}~\cite{JoseEnriquePons2012}.

\begin{definition}
A \textbf{Fuzzy Validity Period} (\emph{FVP}) is defined as a fuzzy time interval specifying when the data regarding an object are valid. A fuzzy time interval is then the fuzzification of a crisp time interval.\\
\end{definition}

Several options to transform possibility distributions corresponding to the fuzzy starting point and the fuzzy end point into one consistent FVP exist \cite{Garrido2009}, e.g (Fig. \ref{fig:fuzzy-validity-period}):
\begin{itemize}
\item The \textbf{convex hull} approach is the most intuitive approach. The resulting FVP is the convex hull of the union of both possibility distributions.
\item The \textbf{uncertainty preserving} approach. The amount of uncertainty is maintained at the edges of the possibility distribution representing the FVP \cite{Garrido2009}.
\end{itemize}

%%%%%%
% FUZZY VALIDITY PERIOD
%%%%%%
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.25]{./graphs/comparisoncv.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:fuzzy-validity-period}Transformation to obtain the FVP. The top graph shows the two triangular possibility distributions. The middle graph shows the convex hull validity period, the bottom one shows the result of the second transformation, which maintains the imprecision.}
%  \label{fig:fuzzy-validity-period}
\vspace*{13pt}

@@

\begin{definition}
A \textbf{Possibilistic Valid-time Period} (\emph{PVP}) is an ill-known interval of time specifying when the data regarding an object are valid.
\end{definition}
A PVP is an ill-known interval in the sense of definition \ref{def;possibilistic-variable}. Note that this representation is \emph{disjunctive}: the PVP represents only one crisp time interval, but for some reason, it is (partially) unknown.

@@

%%
%% comparison table between pvp and fvp.
%%
\vglue13pt
%\begin{table}[htbp]
\tcap{Comparative PVP vs FVP}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{2cm} p{2cm}}\\
Item & PVP & FVP\\
\hline
(1) & $\Pow(\R)$ & $\R$ \\
(2) & Ill-known constraints. & Ad-hoc operators. \\
(3) & $\checkmark$ & - \\
(4) & Two distributions one for each endpoint. & Only one distribution. \\ 
(5) & $\checkmark$ & $\checkmark$ \\
(6) & $\checkmark$ & - \\
\hline\\
\end{tabular}
\label{tbl:comparative-pvp-fvp}
}

In the rest of the paper we will work only with PVP to represent valid-time intervals.

\subsection{\label{subsubsec:Understanding-valid-time-databases}Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the sake of simplicity, only the three main operations in the Data Manipulation Language (CReate Update, Delete) are shown. Usually the DML operations in a temporal database are re-defined (a typical update sentence in SQL could be re-defined by means of several insert and update sentences). Therefore, for the sake of clarity, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally, an illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

\begin{definition}
\emph{Valid-time relation.}
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval.  The pair $\left(A, I\right)$ expresses that the data regarding the entities A are valid during the crisp time interval I. Let $R \subseteq A\  x\  I$ be a valid-time relation. Then, the following equation indicates that the pair $\left(A, I\right)$ is in the relation R:

@@

\begin{definition}
 \label{def:from-beginning}
\emph{From the beginning (FB).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The from the beginning constant is used when the value for $s_i$ is equal to $-\infty$.
\end{definition}

\begin{definition}
 \label{def:until-changed}
\emph{Until changed (UC).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The until changed constant is used when the value for $e_i$ is equal to $+\infty$.
\end{definition}

If the pair $\left(A, I \right)$ with $I = \left(c, UC \right), c \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the patient with ID = 3 was in the hospital the  4/4/2012 and it is still there. The patient with ID = 3 is current in the relation.

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}

\begin{definition}
\label{def:current-in-relation}
\emph{Current $\left(R, A \right)$.}
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,UC \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The function Current $\left(R, A \right)$ returns the crisp time interval $I$ if the entity $A$ is currently in the relation $R$ or the empty set, if the entity $A$ is not currently in the relation.

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \mbox{ and } I = \left(c, UC \right), c \in \T\\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`John')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. On the other hand, Current(R, (4,`Antoon')) returns the empty set.

@@

\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.5]{./graphs/allen.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:allen-rel}Allen's relations for two crisp time intervals $I$ and $J$. Note that the crisp time interval $I$ is fixed and the different positions of the interval $J$ illustrate the Allen's relations.}
\vspace*{13pt}

\begin{definition}
\emph{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(c > e  \right) \wedge \left(d < f  \right) 
\end{equation}
\end{definition}

\begin{definition}
\emph{During:}
 \label{def:during}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(c < e  \right) \wedge \left(d > e  \right) \wedge \left(d < f \right)
\end{equation}
\end{definition}

@@

\begin{definition}
\emph{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[c ,d \right]$ and $J= \left[e ,f \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[c, e \right[ & \mbox{ if } d = UC \mbox{ and } J \mbox{ During } I \\
I & \mbox{ in any other case.}
\end{cases}
\end{align}
\end{definition}

@@

\begin{definition}
\label{def:close-current-version}
\emph{Close-current$\left(R, \left(A, J  \right)\right)$}.
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $J = \left(s_j,e_j \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J\right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if  Current } \left(R, A \right) \neq \emptyset\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

@@

@@

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$. 
% In the example, this correspond with the following insertion sentences:
% \begin{verbatim}
% Insert(3,'Dracula',
% (15/3/2012, 30/3/2012));
% Insert(4,'Frankenstein',
% (25/3/2012, 4/4/2012));
% Insert(5,'Harry Potter',
% (18/3/2012, 2/4/2012));
% \end{verbatim}

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$. For example, consider that the book with ID 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}.
%  The insert sentence is:
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (4/4/2012, UC));
% 	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the book with ID 3 is loaned  the 24/4/2012. Here the problem is that the book with ID 3 was loaned on 4/4/2012, but not returned. If the book is already in the library for a new loan, then it is necessary to set the return date and add a new row with the new loan date. This is illustrated in Table \ref{tbl:library-sample-2}.
%  the insert sentence is:
% 	    \begin{verbatim}
% Insert(3,'Dracula',
% (24/4/2012, UC));
% 	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the object $A$ in the relation $R$. For example, consider now that the librarian wants to introduce a past loan for the book with ID 3. The book was loaned on 6/4/12/2012 and returned on 25/4/2012. As this interval does overlaps other time intervals, it is not possible that the book was loaned in that time interval.Therefore, the insertion is rejected.
%  The insert sentence is:
% 
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (6/4/12/2012,25/4/2012));
% 	    \end{verbatim}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample library database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan Potter & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & 23/4/2012 \\
3 & John & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}

The algorithm for the implementation of the insert operation is the following:

@@

\subsubsection{\label{subsubsec:del}Delete}
The delete operation logically removes a current entity $A$ which is valid in the relation $R$:
\begin{align}
\label{eq:delete}
\DEL \left(R ,A \right) =\\
\begin{cases}
\nonumber
R - \left \lbrace \left(A, I_1 \right),  \ldots, \left(A, I_1 \right)   \right \rbrace, & \forall I_i : \left( A, I_i \right) \in R\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

For example, consider that the hospital wants to delete the history for the patient with ID = 3. The result of this operation is shown in Table \ref{tbl:library-sample-3}.

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-3}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave\\
\hline
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
\hline\\
\end{tabular}
}

>>>>

In this paper we will consider two kinds of imperfection:
\begin{itemize}
\item \textbf{Imperfection in the database}; the knowledge about the temporal data contains some imperfection. E.g., a database record shows that \emph{`The car was in the garage around April.'}
 \item \textbf{Imprecision in the query specification}; it denotes the imprecision in the specification of temporal criteria by the user, when querying. E.g., \emph{`The user wants a car which was in the garage around April.'}
\end{itemize}

@@

In order to deal with uncertainty in time intervals, several proposals have been made. Here, two approaches are described: the first one, based on \emph{Fuzzy Validity Periods}~\cite{Garrido2009} and the second one, based on \emph{Possibilistic Valid-time Periods}~\cite{JoseEnriquePons2012}.

\begin{definition}
A \textbf{Fuzzy Validity Period}\cite{Garrido2009} (\emph{FVP}) is defined as a fuzzy time interval specifying when the data regarding an object are valid. A fuzzy time interval is then the fuzzification of a crisp time interval.\\
\end{definition}

Several options to transform possibility distributions corresponding to the fuzzy starting point and the fuzzy end point into a consistent FVP exist \cite{Garrido2009}, e.g (Fig. \ref{fig:fuzzy-validity-period}):
\begin{itemize}
\item The \textbf{convex hull} approach is the most intuitive approach. The resulting FVP is the convex hull of the union of both possibility distributions.
\item The \textbf{uncertainty preserving} approach. The amount of uncertainty is maintained at the edges of the possibility distribution representing the FVP \cite{Garrido2009}.
\end{itemize}

%%%%%%
% FUZZY VALIDITY PERIOD
%%%%%%
\begin{samepage}
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.25]{./graphs/comparisoncv.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\end{samepage}

\fcaption{\label{fig:fuzzy-validity-period}Transformation to obtain the FVP. The top graph shows the two triangular possibility distributions. The middle graph shows the convex hull validity period, the bottom one shows the result of the second transformation, which maintains the imprecision.}
%  \label{fig:fuzzy-validity-period}
\vspace*{13pt}

@@

\begin{definition}
A \textbf{Possibilistic Valid-time Period} (\emph{PVP}) is an ill-known interval of time specifying when the data regarding an object are valid.
\end{definition}
A PVP is an ill-known interval in the sense of definition \ref{def;possibilistic-variable}. Note that the PVP represents only one crisp time interval, but for some reason, it is (partially) unknown.

@@

%%
%% comparison table between pvp and fvp.
%%
\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{Comparative PVP vs FVP}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{2cm} p{2cm}}\\
Item & PVP & FVP\\
\hline
(1) & $\Pow(\R)$ & $\R$ \\
(2) & Ill-known constraints. & Ad-hoc operators. \\
(3) & $\checkmark$ & - \\
(4) & Two distributions one for each endpoint. & Only one distribution. \\ 
(5) & $\checkmark$ & $\checkmark$ \\
(6) & $\checkmark$ & - \\
\hline\\
\end{tabular}
\label{tbl:comparative-pvp-fvp}
} 
\end{samepage}
In the rest of the paper we will work only with PVP to represent valid-time intervals.

\subsection{\label{subsubsec:Understanding-valid-time-databases}Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the sake of simplicity, only the three main operations in the Data Manipulation Language (CReate Update, Delete) are shown. Usually the DML operations in a temporal database are re-defined (a typical update sentence in SQL could be expressed by means of a couple of insert and update sentences). Therefore, for the sake of clarity, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally, an illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

\begin{definition}
\emph{Valid-time relation.}
Consider $\A$ be the set of all the entities, and let $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity. Let $I = \left(S_i, E_i \right)$ be the set of attributes defining the time interval. The set $\I$ is the set of all the time intervals. The pair $\left(A, I\right)$ expresses that the data regarding the entity A are valid during the crisp time interval I. Let $R \subseteq \A \  x\  \I$ be a valid-time relation. Then, the following equation indicates that the pair $\left(A, I\right)$ is in the relation R:

@@

\begin{definition}
 \label{def:from-beginning}
\emph{From the beginning (FB).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say $I = \left(FB, E_i \right)$ when $I = \left(- \infty, E_i \right)$.
%The from the beginning constant is used when the value for $s_i$ is equal to $-\infty$.
\end{definition}

\begin{definition}
 \label{def:until-changed}
\emph{Until changed (UC).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity. Let $\T$ be the time domain, and $I = \left(S_i, E_i \right), S_i, E_i \in \T$ the set of attributes defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. We will say  $I = \left(S_i, UC \right)$ when $I = \left(S_i, +\infty \right)$.
%The until changed constant is used when the value for $e_i$ is equal to $+\infty$.
\end{definition}

If the pair $\left(A, I \right)$ with $I = \left(S_i, UC \right), S_i \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, let us consider the last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the patient with ID = 3 was in the hospital the  4/4/2012 and he or she is still there. The patient with ID = 3 is current in the relation.

@@

\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
\label{def:current-in-relation}
\emph{Current $\left(R, A \right)$.}
Consider $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i, UC \right)$ the set of attribute defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. The function Current $\left(R, A \right)$ returns the crisp time interval $I$ if the entity $A$ is currently in the relation $R$ or the empty set, if the entity $A$ is not currently in the relation. It is defined as follows:

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \mbox{ and } I = \left(S_i, UC \right), S_i \in \T\\
\emptyset & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`John')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. Conversely, Current(R, (4,`Antoon')) returns the empty set.

@@

\begin{samepage}
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.5]{./graphs/allen.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:allen-rel}Allen's relations for two crisp time intervals $I$ and $J$. Note that the crisp time interval $I$ is fixed and the different positions of the interval $J$ illustrate the Allen's relations.}
\vspace*{13pt}
\end{samepage}

We will use two of the Allen's relations: Overlaps and During which will be defined as follows.

\begin{definition}
\emph{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(S_i, E_i \right)$ and $J = \left(S_j, E_j\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(S_i > S_j  \right) \wedge \left(E_i < E_j  \right) 
\end{equation}
\end{definition}

\begin{definition}
\emph{During:}
 \label{def:during}
Given two time intervals $I = \left(S_i, E_i \right)$ and $J = \left(S_j, E_j\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(S_i < S_j  \right) \wedge \left(E_i > S_j  \right) \wedge \left(E_i < E_j \right)
\end{equation}
\end{definition}

@@

\begin{definition}
\emph{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[S_i ,E_i \right]$ and $J= \left[S_j ,E_j \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[S_i, S_j \right[ & \mbox{ if } E_i = UC \mbox{ and } J \mbox{ During } I \\
I & \mbox{ otherwise}
\end{cases}
\end{align}
\end{definition}

@@

\begin{definition}
\label{def:close-current-version}
\emph{Close-current$\left(R, \left(A, J  \right)\right)$}.
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $J = \left(S_j,E_j \right)$ the set of attribute defining the time interval. Let $R \subseteq \A\  x\  \I$ be a valid-time relation. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J\right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if  Current } \left(R, A \right) \neq \emptyset\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ otherwise}
\end{cases}
\end{eqnarray}
\end{definition}

@@

\begin{definition}
 \emph{Modify(R, (A,I))}.
Let $R \subseteq \A\  x\  \I$ be a valid-time relation and  $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i,E_i\right)$ the set of attributes defining the time interval. The algorithm for the modify operation is defined as follows.
\end{definition}

@@

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an entity $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The entity $A$ was never in the relation $R$: The entity is added with the valid-time indicated by the crisp interval $I$. 
% In the example, this correspond with the following insertion sentences:
% \begin{verbatim}
% Insert(3,'Dracula',
% (15/3/2012, 30/3/2012));
% Insert(4,'Frankenstein',
% (25/3/2012, 4/4/2012));
% Insert(5,'Harry Potter',
% (18/3/2012, 2/4/2012));
% \end{verbatim}

\item The entity $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$ for the entity $A$. For example, consider that the book with ID = 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}.
%  The insert sentence is:
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (4/4/2012, UC));
% 	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the patient with ID = 3 arrived to the hospital on 24/4/2012. Here the problem is that the patient with ID = 3 arrived on 4/4/2012 but, for some reason, the departure date was not stored. If the patient is already at the hospital, then it is necessary to set the departure date and add a new row with the new arrival date. This is illustrated in Table \ref{tbl:library-sample-2}.
%  the insert sentence is:
% 	    \begin{verbatim}
% Insert(3,'Dracula',
% (24/4/2012, UC));
% 	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the entity $A$ in the relation $R$. For example, consider that the hospital manager wants to introduce a past arrival for the patient with ID = 3. The patient arrived on 6/4/2012 and left the hospital on 25/4/2012. As the dates do overlap, it is not possible that the patient was in the hospital at that time interval. Therefore, the insertion is rejected.
%  The insert sentence is:
% 
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (6/4/12/2012,25/4/2012));
% 	    \end{verbatim}

@@

\end{enumerate}
\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan Potter & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & 23/4/2012 \\
3 & John & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}
\end{samepage}

\begin{definition}
 \emph{Insert(R,(A,I)))}.
Let $R \subseteq \A\  x\  \I$ be a valid-time relation and  $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i,E_i\right)$ the set of attributes defining the time interval. Then, the algorithm for the implementation of the insert operation is defined as follows.
\end{definition}

@@

\subsubsection{\label{subsubsec:del}Delete}
The delete operation logically removes a current entity $A$ which is valid in the relation $R$.
\begin{definition}
 \emph{Delete(R,(A,I))}.
Let $R \subseteq \A\  x\  \I$ be a valid-time relation and  $A=\left(A_1, \ldots, A_n \right), A \in \A$ be the set of attributes that define an entity and $I = \left(S_i,E_i\right)$ the set of attributes defining the time interval. The algorithm for the delete operation is defined as follows.
\end{definition}

\begin{align}
\label{eq:delete}
\DEL \left(R ,A \right) =\\
\begin{cases}
\nonumber
R - \left \lbrace \left(A, I_1 \right),  \ldots, \left(A, I_n \right)   \right \rbrace, & \forall I_i : \left( A, I_i \right) \in R\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

For example, consider that the hospital manager wants to delete the history for the patient with ID = 3. The result of this operation is shown in Table \ref{tbl:library-sample-3}.

@@

\begin{samepage}
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-3}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave\\
\hline
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
\hline\\
\end{tabular}
}
\end{samepage}
